---
title: "Exploratory Data Analysis"
author: "Rebecca Chang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading Necessary Packages
```{r}
library(tidyverse)
library(sf)

```
#Reading In Data/Merging Datasets
```{r}
PersonData <- read_rds('../data/PersonData_111A.Rds')
HHData <- read_rds('../data/HHData_111A.Rds')
hh_bgDensity <- read_rds('../data/hh_bgDensity.Rds')

personHHData <- left_join(PersonData, HHData) %>% left_join(hh_bgDensity)

county_shp <- st_read("../data/counties/counties.shp")
```
#Quick Plot
```{r}
plot(county_shp["NAME"])
```
The unique colors correspond to the unique names of each county. However, you will notice that there are repeat colors because the color palette available when plotting in such a simple way is limited.


#List Variable Names
```{r}
names(personHHData)
```
#We will now be taking some of the CHTS person and household characteristics and aggregating them to the county level so we can map the traits of our CHTS respondents county-by-county.

#To start off, let’s get the total count of how many people in our survey come from the various counties of California:
```{r}
personHHData %>% 
  group_by(CTFIP, County) %>% #added County to this grouping so we can see the county names
  summarise(count=n()) %>% 
  view()
```

```{r}
prhh_aggreg <- personHHData %>% 
  group_by(County, CTFIP) %>%
  mutate(count = n()) %>% # the new column called 'count'
  summarise_at(vars(-hhid, -pnum, -bg_group), mean)

View(prhh_aggreg)
```

```{r}
county_prhh_shp <- county_shp %>% left_join(prhh_aggreg)
view(county_prhh_shp) # check the data to see what we get
```


# Map of the percentage of CHTS respondents born in the USA per county.

```{r}
mapview(county_prhh_shp, # the dataset to use 
        zcol = "bornUSA", # tells it which column to map
        legend = TRUE, # if FALSE, it won't show the legend 
        label = as.character(county_prhh_shp$NAME), # tells it the column whose value you want to appear when you hover over a shape with your mouse
        popup = leafpop::popupTable(x = county_prhh_shp, zcol = c("bornUSA", "count"))  # determines what is included in the popup window when you click on a shape
        )
```

# Group the data we have into 4 parts based on their bg_group
```{r}
county_bg_aggreg <- personHHData %>% 
  group_by(County, CTFIP, bg_group) %>%  # group by county, CTFIP, and also bg_group
  mutate(count = n()) %>% 
  summarise_at(vars(-hhid, -pnum), mean)

view(county_bg_aggreg)

county_bg_shp <- county_shp %>% 
  merge(data.frame(bg_group = c("Urban", "Suburban", "Exurban", "Rural"))) %>% 
  left_join(county_bg_aggreg)
```


```{r}
urban_TripMap <-  mapview(filter(county_bg_shp, bg_group == "Urban"),
                          zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                          layer.name = "Urban Trips")

suburb_TripMap <- mapview(filter(county_bg_shp, bg_group == "Suburban"),
                          zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                          layer.name = "Suburban Trips")

exurb_TripMap <- mapview(filter(county_bg_shp, bg_group == "Exurban"),
                         zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                         layer.name = "Exurban Trips")

rural_TripMap <- mapview(filter(county_bg_shp, bg_group == "Rural"),
                         zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                         layer.name = "Rural Trips")
```



```{r}
mapview(county_prhh_shp, 
        zcol = "Sum_PMT", 
        legend = TRUE, 
        label = as.character(county_prhh_shp$CTFIP), 
        popup = leafpop::popupTable(x = county_prhh_shp, 
                                    zcol = c("NAME", "count", "Sum_PMT"))  
)
```

