---
title: "Exploratory Data Analysis"
author: "Rebecca Chang"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Necessary Packages

```{r}
library(tidyverse)

# Load spatial data
require(sf) || install.packages("sf", dependencies = TRUE)
library(sf)

require(mapview) || install.packages("mapview", dependencies = TRUE)
library(mapview)
mapviewOptions(fgb = FALSE)

require(leafsync) || install.packages("leafsync", dependencies = TRUE)
library(leafsync)

require(maps) || install.packages("maps", dependencies = TRUE)
library(maps)
```

## Combine Person/Household Data

```{r}
PersonData <- read_rds('../data/PersonData_111A.Rds')
HHData <- read_rds('../data/HHData_111A.Rds')
hh_bgDensity <- read_rds('../data/hh_bgDensity.Rds')

personHHData <- left_join(PersonData, HHData) %>%
  left_join(hh_bgDensity)

county_shp <- st_read("../data/counties/counties.shp")
```

## Quick Plot

```{r}
plot(county_shp["NAME"])  
```

The unique colors correspond to the unique names of each county.

## List Variable Names

```{r}
names(personHHData)
```

# Interactive Map

## Aggregating CHTS traits to county level and joining to shapefile

Taking some of the CHTS person and household characteristics and aggregating them to the county level so we can map the traits of our CHTS respondents county-by-county.

Get the total count of how many people in our survey come from the various counties of California:

```{r}
personHHData %>% 
  group_by(CTFIP, County) %>% # added County to this grouping so we can see the county names
  summarise(count=n())
```

We get the county-by-county means of all the variables of interest. Taking the means instead of just getting sums accounts for the fact that there are not equal amounts of people surveyed from each county. We also create a new column `count` to this dataset so we can carry over the counts of how many people were surveyed from each county.

```{r}
prhh_aggreg <- personHHData %>% 
  group_by(County, CTFIP) %>%
  mutate(count = n()) %>% # the new column called 'count'
  summarise_at(vars(-hhid, -pnum, -bg_group), mean)
```

## Join datasets on CTFIP (county ID) variable

```{r}
county_prhh_shp <- county_shp %>% left_join(prhh_aggreg)
```

Newly-created `county_prhh_shp` is also an `sf` object so it has spatial attributes and can be mapped.

## Map of the percentage of CHTS respondents born in the USA per county.

```{r}
mapview(county_prhh_shp, # the dataset to use 
        zcol = "bornUSA", # tells it which column to map
        legend = TRUE, # if FALSE, it won't show the legend 
        label = as.character(county_prhh_shp$NAME), # tells it the column whose value you want to appear when you hover over a shape with your mouse
        popup = leafpop::popupTable(x = county_prhh_shp, zcol = c("bornUSA", "count"))  # determines what is included in the popup window when you click on a shape
        )
```

Notice that the percentage of respondents who were born in the USA increases as you look at more northern counties. However, also notice that more urban counties have lower percentages, regardless of their position in the state.

# Lattice of Interactive Maps

## Group the data we have into 4 parts based on their bg_group

4 maps for each residential location type

```{r}
county_bg_aggreg <- personHHData %>% 
  group_by(County, CTFIP, bg_group) %>%  # group by county, CTFIP, and also bg_group
  mutate(count = n()) %>% 
  summarise_at(vars(-hhid, -pnum), mean)

county_bg_shp <- county_shp %>% 
  merge(data.frame(bg_group = c("Urban", "Suburban", "Exurban", "Rural"))) %>% 
  left_join(county_bg_aggreg)
```

Sum Trips by Residential Area

```{r}
urban_TripMap <-  mapview(filter(county_bg_shp, bg_group == "Urban"),
                          zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                          layer.name = "Urban Trips")

suburb_TripMap <- mapview(filter(county_bg_shp, bg_group == "Suburban"),
                          zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                          layer.name = "Suburban Trips")

exurb_TripMap <- mapview(filter(county_bg_shp, bg_group == "Exurban"),
                         zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                         layer.name = "Exurban Trips")

rural_TripMap <- mapview(filter(county_bg_shp, bg_group == "Rural"),
                         zcol = "Sum_Trips", legend = TRUE, popup = NULL,
                         layer.name = "Rural Trips")
```

```{r}
latticeview(urban_TripMap, suburb_TripMap, exurb_TripMap, rural_TripMap, sync = "all")
```

Sum PMT/Distance by Residential Area

```{r}
urban_PMTMap <- mapview(filter(county_bg_shp, bg_group == "Urban"),
                        zcol = "Sum_PMT", legend = TRUE, popup = NULL,
                        layer.name = "Urban PMT")

suburb_PMTMap <- mapview(filter(county_bg_shp, bg_group == "Suburban"),
                         zcol = "Sum_PMT", legend = TRUE, popup = NULL,
                         layer.name = "Suburban PMT")

exurb_PMTMap <- mapview(filter(county_bg_shp, bg_group == "Exurban"),
                        zcol = "Sum_PMT", legend = TRUE, popup = NULL,
                        layer.name = "Exurban PMT")

rural_PMTMap <- mapview(filter(county_bg_shp, bg_group == "Rural"),
                        zcol = "Sum_PMT", legend = TRUE, popup = NULL,
                        layer.name = "Rural PMT")
```

```{r}
latticeview(urban_PMTMap, suburb_PMTMap, exurb_PMTMap, rural_PMTMap, sync = "all")
```

# Static Map

```{r}
county <- ggplot2::map_data("county", region = "california") # get the CA county data

county_bg <- merge(county, data.frame(bg_group = c("Urban", "Suburban", "Exurban", "Rural")))

county_bg_all <- county_bg_aggreg %>% 
  mutate(subregion = tolower(County)) %>% 
  full_join(county_bg, by = c("subregion", "bg_group"))
```

```{r}
ggplot(county_bg_all) +
  geom_polygon(aes(x = long, y = lat, group = subregion, fill = Sum_PMT), colour = "white") +
  scale_fill_distiller(palette = "YlGnBu", direction = 1) +
  facet_wrap(vars(bg_group), nrow = 2) +  # multi-panel plots using facet_wrap(), plot in 2 rows
  ggtitle("Total PMT in California at County-level") + 
  theme_void() +
  theme(legend.position="bottom")
```
